GEE models

### 2017-10-17 doc/manuscript.Rmd
minor typos/edits before sending to Tony

- AR1: Autoregressive of order 1
- ALT: Alanine aminotransferase
- EPA: Eicosapentaenoic acid
- FDR: False Discovery Rate
- GEE: Generalized estimating equations
- HOMA: Homeostatic model assessment
- HOMA-IR: Homeostatic model assessment of insulin resistance
- HOMA2-%S: Homeostatic model assessment 2 of insulin sensitivity
- IFG: Impaired fasting glucose
- IGI/IR: Insulinogenic index over HOMA-IR
- IGT: Impaired glucose tolerance
- ISI: Insulin Sensitivity Index (Matsuda index)
- ISSI-2: Insulin Secretion-Sensitivity Index-2
- MAQ: Modified Activity Questionnaire
- MET: Metabolic Equivalent of Task
- NEFA: Non-esterified fatty acids
- OGTT: Oral glucose tolerance test
- PLS: Partial least squares regression
- PROMISE: Prospective Metabolism and Islet Cell Evaluation Cohort
- PUFA: Polyunsaturated fatty acids
- QIC: Quasi-Likelihood Information Criterion
- TAG: Triacylglyceride
- WC: Waist circumference

@@

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Roche Canada, Mississauga, Ontario, Canada).  Both insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyser and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc, St. Charles, Missouri, USA), and has a coefficient of variation (CV) of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics, Mississauga, Ontario, Canada) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorised using the 2006 WHO criteria [@WHO2006].

@@

Insulin sensitivity and beta cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
and the Insulin Sensitivity Index (ISI) [@Matsuda1999]. HOMA2-%S largely
reflects hepatic insulin sensitivity, while ISI reflects whole-body insulin
sensitivity [@AbdulGhani2007].  beta cell function was assessed using the
Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR is a measure of
the first phase of insulin secretion while ISSI-2 is analogous to the
disposition index (but using OGTT values). Each index has been validated against
gold standard measures [@Hermans1999a;@Matsuda1999;@Retnakaran2009]. Formulas
used to calculate these measures can be found in the ESM Methods.

@@

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` kg/m^2^ (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from GEE analyses (n={{num}} at 3rd year, n=
{{num}} at 6th year). Follow-up data were included until date of
diagnosis.
Certain medications (e.g. lipid medications such as statins or
fibrates) could influence the association between NEFA and the outcomes.
There were `r meds_use[1]` and `r meds_use[2]` participants who used lipid and
cholesterol modifying medications at the baseline and 6-year visit,
respectively. Use of these medications did not influence NEFA composition, the
outcomes, or the GEE models (data not shown).

>>>>

- ALT: Alanine aminotransferase
- AR1: Autoregressive of order 1
- EPA: Eicosapentaenoic acid
- FDR: False Discovery Rate
- GEE: Generalized estimating equations
- HOMA: Homeostatic model assessment
- HOMA-IR: Homeostatic model assessment of insulin resistance
- HOMA2-%S: Homeostatic model assessment 2 of insulin sensitivity
- IFG: Impaired fasting glucose
- IGI/IR: Insulinogenic index over HOMA-IR
- IGT: Impaired glucose tolerance
- ISI: Insulin Sensitivity Index (Matsuda index)
- ISSI-2: Insulin Secretion-Sensitivity Index-2
- MAQ: Modified Activity Questionnaire
- MET: Metabolic Equivalent of Task
- NEFA: Non-esterified fatty acids
- OGTT: Oral glucose tolerance test
- PLS: Partial least squares regression
- PROMISE: Prospective Metabolism and Islet Cell Evaluation Cohort
- PUFA: Polyunsaturated fatty acids
- QIC: Quasi-Likelihood Information Criterion
- TAG: Triacylglyceride
- WC: Waist circumference

@@

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Roche Canada, Mississauga, Ontario, Canada).  Both insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyser and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc, St. Charles, Missouri, USA), and has a coefficient of variation (CV) of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics, Mississauga, Ontario, Canada) with a detection range of 0.11 (2 mg/dl) to 41.6 mmol/l.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorised using the 2006 WHO criteria [@WHO2006].

@@

Insulin sensitivity and beta cell function indices were computed using the OGTT 
glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
and the Insulin Sensitivity Index (ISI) [@Matsuda1999]. HOMA2-%S largely
reflects hepatic insulin sensitivity, while ISI reflects whole-body insulin
sensitivity [@AbdulGhani2007].  Beta cell function was assessed using the
Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR is a measure of
the first phase of insulin secretion while ISSI-2 is analogous to the
disposition index (but using OGTT values). Each index has been validated against
gold standard measures [@Hermans1999a;@Matsuda1999;@Retnakaran2009]. Formulas
used to calculate these measures can be found in the ESM Methods.

@@

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` kg/m^2^ (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")` incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from GEE analyses (n={{num}} at 3rd year, n=
{{num}} at 6th year). Follow-up data were included until date of
diagnosis.
Certain medications (e.g. lipid medications such as statins or
fibrates) could influence the association between NEFA and the outcomes.
There were `r meds_use[1]` and `r meds_use[2]` participants who used lipid and
cholesterol modifying medications at the baseline and 6-year visit,
respectively. Use of these medications did not influence NEFA composition, the
outcomes, or the GEE models (data not shown).
